Large and Infinitary Quotient Inductive-Inductive Types

### 2020-01-08 draft/paper.tex
paper

\newcommand{\arri}{\Rightarrow}
\newcommand{\arre}{\Rightarrow^{\mathsf{ext}}}
\newcommand{\arrinf}{\Rightarrow^{\mathsf{inf}}}

@@

%% Keywords
%% comma separated list
\keywords{inductive types, quotient inductive-inductive types}  %% \keywords are mandatory in final camera-ready submission

@@

\todo{revise metatheory intro, use Timany cumulativity}
\todo{recheck universe levels, prettify levels}
\todo{explain the limitations and design choices in univ levels}
\todo{write the rest}
\todo{see if we can cut back on flCwF, add more examples instead}

@@

Categories are a good example. Signatures for QIITs allow having multiple sorts,
with later ones indexed over previous ones, and equations as well. We need both
features in order to write down the signature of categories.

@@

Algebraic notions of models of type theories are examples for this. Here,
initial models represent syntax, and initiality corresponds to induction on
syntax. Several variants have been used, from contextual categories \cite{TODO}
and comprehension categories \cite{TODO} to categories with families
\cite{TODO}.

Real, surreal \cite{TODO} and ordinal numbers \cite{TODO} are examples of
\emph{infinitary} QIITs, which are represented as inductive trees which may
branch infinitely.

@@

\subsection{Contributions}
The current paper extends previously formalized QIITs
\cite{kaposi2019constructing}. We add the following features:
\begin{enumerate}
  \item
  \textbf{Large constructors, large elimination} and algebras at different
  universe levels. This fills in an important formal gap, since large
  elimination is routinely used in the metatheory of type theories, but it has
  not been but it has not been presented explicitly in previous works about
  QIITs.
  \item
  \textbf{Infinitary constructors}.
  Of special note here is that the theory of QIIT signatures is itself large and
  infinitary, thus it can ``eat itself'', i.e.\ include its own signature and
  provide its own metatheory. This was not possible previously in
  \cite{kaposi2019constructing}, where only finitary QIITs were described. In
  this paper we use self-representation to bootstrap the model theory of
  signatures, without having to assume any pre-existing internal syntax.
  \item
  \textbf{Recursive equations}, i.e. equations appearing as assumptions
  of constructors. These have occurred previously in syntaxes of cubical type
  theories, as boundary conditions\cite{TODO}.
\end{enumerate}

@@

Compared to the case of finitary QIITs, the addition of infinitary constructors
and recursive equations requires a significant change in semantics: instead of
strict CwF morphisms, we need to consider weak ones, and instead of modeling
types as displayed CwFs, we need to model them as CwF isofibrations. In a
nutshell, the new semantics must be given mutually with a form of a ``structure
identity principle''\cite{TODO}, which says that signature extension respects algebra
isomorphisms.

@@

In Section \ref{sec:metatheory}, we describe the metatheory used in the rest of
the paper. In Section \ref{sec:tos}, we introduce the theory of QIIT
signatures. In Section \ref{sec:semantics} we give categorical semantics to
signatures. In Section \ref{sec:tossig} we build model theory for the theory
of QIIT signatures. In Section \ref{sec:termmodels} we give a term model construction
of QIITs. We discuss related work and conclude in Sections \ref{TODO}

@@

Both for functions and $\Sigma$, the output universe level is given as the
maximum of the levels of the constituent types, e.g.\ $(x : A)\ra B : \Set_{\max(i,j)}$
when $A : \Set_i$ and $B : \Set_j$.

@@

\subsection{Cumulativity}

@@

A \textbf{category with families} (CwF), where all four underlying sets (of
objects, morphisms, types and terms) are in $\Set_j$. Following notation in
\cite{kaposi2019constructing}, we denote these respectively as $\Con : \Set_j$,
$\Sub : \Con \ra \Con \ra \Set_j$, $\Ty : \Con \ra \Set_j$ and $\Tm : (\Gamma :
\Con) \ra \Ty\,\Gamma \ra \Set_j$. We use $\id$ and $\blank\circ\blank$ to
denote identity and composition for substitution. We denote the empty context as
$\emptycon : \Con$, and the unique substitution into the empty context as
$\epsilon : \Sub\,\Gamma\,\emptycon$. Context extension is $\blank\ext\blank :
(\Gamma : \Con)\ra \Ty\,\Gamma \ra \Con$. Substitution on types and terms is
written as $\blank[\blank]$. Projections are noted as $\p :
\Sub\,(\Gamma\,\ext\,A)\,\Gamma$ and $\q : \Tm\,(\Gamma\,\ext\,A)\,(A[\p])$, and
substitution extension is $\blank,\blank : (\sigma : \Sub\,\Gamma\,\Delta)\ra
\Tm\,\Gamma\,(A[\sigma])\ra \Sub\,\Gamma\,(\Delta\ext A)$.

@@

\textbf{External function space} $\Pie : (A : \Set_i)\ra(A \ra \Ty\,\Gamma)\ra
\Ty\,\Gamma$, with $\appe : \Tm\,\Gamma\,(\Pie\,A\,B)\ra((x : A)\ra
\Tm\,\Gamma\,(B\,x))$ and its inverse $\lame$.

\textbf{Infinitary function space} $\Piinf : (A : \Set_i)\ra(A \ra
\Tm\,\Gamma\,\U)\ra \Tm\,\Gamma\,\U$, with $\appinf :
\Tm\,\Gamma\,(\El\,(\Pie\,A\,b))\ra((x : A)\ra \Tm\,\Gamma\,(\El\,(b\,x)))$ and
its inverse $\laminf$.

@@

\begin{definition}[Notion of signature]\label{def:signature}
A QIIT signature at level $i$ is a context in an arbitrary $M : \ToS_{i,j}$
model. We define the type of such signatures as follows:
\[
  \Sig_i :\equiv (j : \mathbb{N})\ra(M : \ToS_{i,j})\ra \Con_M
\]

Here, $i$ refers to the level of external types appearing in the signature, in
the domains of $\Pie$ and $\Piinf$ functions, while the quantified $j$ level is
required to allow interpreting a signature in arbitrary-sized ToS models. Note
that $\Sig_i$ is universe-polymorphic, so it is a family of cETT types and it is
not in any cETT universe.

\begin{example}{
    Signature for natural numbers. Here, no external types appear, so the level
    can be chosen as $0$.}
\begin{alignat*}{3}
  & NatSig : \Sig_0 \\
  & NatSig :\equiv \lambda (j :\mathbb{N})(M : \ToS_{0,j}).\\
  & \hspace{1em}(\emptycon_M\,\ext_M\, (N : \U_M) \,\ext_M\,(zero : \El_M\,N)\\
  & \hspace{2.67em}\ext_M (suc : N\arri_M\El_M\,N))
\end{alignat*}
\end{example}

With this, we are able to specify QIITs, and we can also interpret each
signature in an arbitrary ToS model, by applying a signature to a model.
$\Sig_i$ can be viewed as a precursor to a Church-encoding for the theory of
signatures, but we only need contexts encoded in this way, and not other ToS
components. In functional programming, this is sometimes called
``finally tagless''\cite{carette2007finally}, and it is used for defining and
interpreting embedded languages.
\end{definition}

@@

\begin{example}{Recursive equations}. Again, the universe is closed under $\Id$, which allows us to write equations
in $\Pii$ domains. A minimal (and trivial) example:
\begin{alignat*}{3}
  & RecEqSig :\equiv\\
  & \hspace{0.5em}\emptycon\,\ext (A : \U) \ext (a : \El\,A)\ext (f : \Pii\,(x : A)\,(\Id\,A\,x\,a\arri\El\,A))
\end{alignat*}
More interesting (and complicated) examples for recursive equations are boundary
conditions in various cubical type theories \cite{TODO}. Note that our $\Id$
allows iterated equations as well, but these are all trivial in the semantics,
where we assume UIP.
\end{example}

@@

In the following, for $i$ and $j$ levels we define a model $\bM_{i,j} : \ToS_{i,
  \max(i,j+1)+1}$, such than $\Con_{\bM_{i,j}}$ is a type of structured categories (of
algebras). The level $i$ marks the level of all external sets in function
domains, and $j$ marks the level of all internal sorts in an algebra. Hence,
every algebra has level $\max(i,j+1)$. The bump is only needed for $j$, since
algebras merely contain elements of $A : \Set_i$ types, while inductive sets are
themselves elements of $\Set_j$. For example, $NatAlg_j : \Set_{\max(0, j+1)} :
\Set_{\max(0, j+1)+1}$,

@@

We define $\bCon : \Set_{\max(i,j+1)+1}$ as $\flCwF_{\max(i,j+1)}$.
%% We define
%% $\flCwF$ as follows.

\begin{definition}[Finite limit CwFs]\label{def:flCwF}
For each level $i$ we define $\flCwF_i : \Set_{1+i}$ as an iterated $\Sigma$-type
with the following components:
\begin{enumerate}
  \item A CwF with underlying sets all in $\Set_i$. We reuse the component
    notations from Definition \ref{def:tos}.
  \item $\Sigma$-type $\Sg : (A : \Ty\,\Gamma)\ra \Ty\,(\Gamma \ext A)
    \ra \Ty\,\Gamma$, with term formers $\proj1$, $\proj2$ and $\blank,\blank$.
  \item Identity type $\Id : (A : \Ty\,\Gamma)\ra \Tm\,\Gamma\,A\ra
    \Tm\,\Gamma\,A\ra \Ty\,\Gamma$, with $\refl$, equality reflection and UIP.
  \item Constant families. This includes a type former $\Kfam : \Con \ra
    \Ty\,\Gamma$, where $\Gamma$ is implicitly quantified, together with $\lamK
    : \Sub\,\Gamma\,\Delta \ra \Tm\,\Gamma\,(\Kfam\,\Delta)$ and its inverse
    $\appK$. The idea is that $\Kfam\,\Delta$ is a representation of $\Delta$ as
    a type in any context. Clairambault and Dybjer called constant families
    ``democracy'' in \cite{clairambault2014biequivalence}.
\end{enumerate}
\end{definition}

@@

\begin{theorem}[Equivalence of initiality and induction]
An object $\Gamma : \Con_{\bGamma}$ supports induction if and only if it is
initial. Moreover, induction and initiality are both proof irrelevant
predicates.
\end{theorem}
\begin{proof} By brief internal flCwF reasoning in \cite{kaposi2019constructing}.
\end{proof}

@@

\subsection{Substitutions}

\begin{definition}\label{def:weakmorphism} A \emph{weak flCwF morphism} $\boldsymbol{\sigma : \Sub\,\Gamma\,\Delta}$ is a functor between underlying categories, which additionally maps types to types and terms to terms, and satisfies the following:
  \begin{enumerate}
    \item $\bsigma\,(A[\sigma]) = (\bsigma\,A)\,[\bsigma\,\sigma]$
    \item $\bsigma\,(t[\sigma]) = (\bsigma\,t)\,[\bsigma\,\sigma]$
    \item The unique map $\epsilon : \Sub\,(\bsigma\,\emptycon)\,\emptycon$ has a retraction.
    \item Each $(\bsigma\,\p,\,\bsigma\,\q) : \Sub\,(\bsigma\,(\Gamma\,\ext\,A))\,(\bsigma\,\Gamma\,\ext\,\bsigma\,A)$ has an inverse.
  \end{enumerate}
\end{definition}

@@

@@

We define $\boldsymbol{\Ty\,\Gamma} : \Set_{1 + \max(i, j+1)}$ as the type of
split flCwF-isofibrations over $\bGamma$. We extend Ahrens' and Lumsdaine's
displayed categories and their definition of isofibrations \cite{displayedcats}.
We first define displayed flCwFs, then specify iso-cleaving as additional structure
on top of that.

\begin{definition}[Displayed flCwF] The type of displayed flCwFs at level $i$ is given as the logical predicate interpretation \cite{TODO} of $\flCwF_i$. For each flCwF component in $\bGamma$, there is a component in a displayed flCwF which ``lies over'' it.

In situations where we need to refer to both ``base'' and displayed things, we give
\ul{underlined} names to contexts, substitutions, types and terms in a base
flCwF. For example, we may have $\ulGamma : \Con_{\bGamma}$ living in
$\boldsymbol{\Gamma : \Con}$, and $\Gamma : \Con_{\bA}\,\ulGamma$ living in
$\boldsymbol{A : \Ty\,\Gamma}$. We only use underlining on cETT variable names,
and overload flCwF component names for displayed counterparts.

@@

In contrast to \cite{kaposi2019constructing}, it is not sufficient to model
types with displayed flCwFs. In ibid.\ substitutions are modeled as strict
morphisms, which we cannot do because our infinitary function types and identity
types cannot be interpreted strictly, as we will see in Section
\ref{TODO}. Strict preservation is expressed with metatheoretic equality, which
is respected by all cETT constructions, but isomorphisms are not necessarily
respected. Hence, we need to add additional structure to displayed flCwFs which
expresses preservation of base isomorphisms.

@@

\emph{Remark.} It is not possible to model types as fibrations or opfibrations, because
we have no restriction on the variance of ToS types. For example, the type which extends
a pointed set to a $Nat$-signature is invariant in the base signature.

@@

\todo{iso-cleaving}

Functoriality of type substitution, i.e.\ $\boldsymbol{A[\id]} = \bA$ and
$\boldsymbol{A[\sigma\circ\delta]} = \boldsymbol{A[\sigma][\delta]}$, follows
from Lemma \ref{lem:idcomppres} and split cleaving given by $\coe_{\id}$,
$\coe_{\circ}$, $\coh_{\id}$ and $\coh_{\circ}$ in $\bA$.

@@

$\boldsymbol{\Tm\,\Gamma\,A} : \Set_{1 + \max(i, 1+j)}$ is defined as the type of
\emph{weak flCwF sections} of $\bA$. The underlying functions of $\bt :
\bTm\,\bGamma\,\bA$ are as follows:
\begin{alignat*}{3}
  & \bt : (\ulGamma : \Con_{\bGamma}) \ra \Con_{\bA}\,\ulGamma\\
  & \bt : (\ulsigma : \Sub_{\bGamma}\,\ulGamma\,\ulDelta)
         \ra \Sub_{\bA}\,(\bt\,\ulGamma)\,(\bt\,\ulDelta)\,\ulsigma\\
  & \bt : (\ulA : \Ty_{\bGamma}) \ra \Ty_{\bA}\,(\bt\,\ulGamma)\,\ulA\\
  & \bt : (\ult : \Tm_{\bGamma}\,\ulGamma\,\ulA) \ra
          \Tm_{\bA}\,(\bt\,\ulGamma)\,(\bt\,\ulA)\,\ult
\end{alignat*}
Such that
\begin{enumerate}
  \item $\bt\,(\ulA[\ulsigma]) = (\bt\,\ulA)\,[\bt\,\ulsigma]$
  \item $\bt\,(\ult[\ulsigma]) = (\bt\,\ult)\,[\bt\,\ulsigma]$
  \item The unique map $\epsilon_{\bA} : \Sub\,(\bt\,\emptycon)\,\emptycon\,\id$ has a vertical retraction.
  \item Each $(\bt\,\p,\,\bt\,\q) : \Sub\,(\bt\,(\ulGamma\,\ext\,\ulA))\,(\bt\,\ulGamma\,\ext\,\bt\,\ulA)\,\id$ has a vertical inverse.
\end{enumerate}

Similarly to Definition \ref{def:weakmorphism}, we denote the evident preservation
isomorphisms as $\bt_{\emptycon} : \bt\,\emptycon \simeq_{\id} \emptycon$ and
$\bt_{\ext} : \bt\,(\ulGamma\ext \ulA) \simeq_{\id} \bt\,\ulGamma \ext
\bt\,\ulA$. In short, weak section is a dependently typed analogue of weak
morphism, with dependent underlying functions and displayed preservation
isomorphisms. We also have the derived fl-preservation isomorphisms.

@@

We define $\boldsymbol{\U : \Ty\,\Gamma}$ as the isofibration which is
constantly $\bSet_j$. A constant isofibration does not actually depend on
the base flCwF, and has trivial iso-cleaving where $\coe$-s are identity
functions. Hence, we have $\Con_{\bU}\,\ulGamma :\equiv \Set_j$ and $\Sub_{\bU}\,\Gamma\,\Delta\,\ulsigma :\equiv \Gamma \ra \Delta$.

\emph{Remark.} The type $\bTm\,\bGamma\,\bU$ is strictly equal to $\bSub\,\bGamma\,\bSet_j$, so in the following we will think about semantic elements of the universe as weak morphisms from $\bGamma$ to $\bSet_j$.

@@

For $\boldsymbol{a : \Tm\,\Gamma\,\U}$ and $\boldsymbol{B :
  \Ty\,(\Gamma\ext\El\,a)}$, we aim to define $\boldsymbol{\Pi\,a\,B}
\boldsymbol{:} \bTy\,\bGamma$. We define this as a dependent product of
isofibrations, indexed by a discrete domain. With a general $\bA \boldsymbol{:}
\bTy\,\bGamma$ domain, $\bPii$ would not be definable, since variance issues
preclude $\Pi$-types in the CwF of categories \cite{TODO}.

%% _{(\bPii\,\ba\,\bB)}

@@

For $A : \Set_i$ and $\bB : A \ra \bTy\,\bGamma$, we define $\bPie\,A\,\bB
\boldsymbol{:} \bTy\,\bGamma$ as the $A$-indexed direct product of $\bB$. Since
the indexing is given by a metatheoretic function, every component is given in the
evident pointwise way.

\subsection{Infinitary Function Space}

>>>>

\newcommand{\arri}{\Rightarrow}
\newcommand{\arre}{\Rightarrow^{\mathsf{ext}}}
\newcommand{\arrinf}{\Rightarrow^{\mathsf{inf}}}
\newcommand{\syn}{\mathsf{syn}}
\newcommand{\SynSig}{\mathsf{SynSig}}

@@

%% Keywords
%% comma separated list
\keywords{inductive types, infinitary inductive types, quotient inductive-inductive types, categories with families}

@@

@@

Categories are a prime example for a structure which is described by a quotient
inductive-inductive signature. Signatures for QIITs allow having multiple sorts,
with later ones indexed over previous ones, and equations as well. We need both
features in order to write down the signature of categories.

@@

Algebraic notions of models of type theories are examples for this. Here,
initial models represent syntax, and initiality corresponds to induction on
syntax. Several variants have been used, from contextual categories \cite{gat}
and comprehension categories \cite{jacobs1993comprehension} to categories with
families \cite{Dybjer96internaltype}. Additionally, real, surreal \cite{HoTTbook}
and ordinal numbers \cite{lumsdaineShulman} can be defined as \emph{infinitary}
QIITs, which are represented as possibly infinitely branching inductive trees.

@@

\subsection{Contributions}
The current paper extends the syntax and semantics of QIITs as previously described
in the literature \cite{kaposi2019constructing, gabe, dijkstra2017quotient}. We add the following features:
\begin{enumerate}
  \item
  \textbf{Large constructors, large elimination} and algebras at different
  universe levels. This fills in an important formal gap; large models are
  routinely used in the metatheory of type theories, but they have not been
  presented explicitly in previous QIIT literature.
  \item
  \textbf{Infinitary constructors}.
  Of special note here is that the theory of QIIT signatures is itself large and
  infinitary, thus it can ``eat itself'', i.e.\ include its own signature and
  provide its own metatheory. This was not possible previously in
  \cite{kaposi2019constructing}, where only finitary QIITs were described. In
  this paper we use self-representation to bootstrap the model theory of
  signatures, without having to assume any pre-existing internal syntax.
  \item
  \textbf{Recursive equations}, i.e. equations appearing as assumptions
  of constructors. These have occurred previously in syntaxes of cubical type
  theories, as boundary conditions\cite{TODO}.
\end{enumerate}

@@

Compared to the case of finitary QIITs, the addition of infinitary constructors
and recursive equations requires a significant change in semantics: instead of
strict CwF morphisms, we need to consider weak ones, and instead of modeling
types as displayed CwFs, we need to model them as CwF isofibrations. The latter
amounts to showing that signature extension respects algebra isomorphisms.

@@

In Section \ref{sec:metatheory}, we describe the metatheory used in the rest of
the paper. In Section \ref{sec:tos}, we introduce the theory of QIIT
signatures. In Section \ref{sec:semantics} we give categorical semantics to
signatures. In Section \ref{sec:tossig} we build model theory for the theory
of QIIT signatures. In Section \ref{sec:termmodels} we give a term model construction
of QIITs. We discuss related work and conclude in Sections \ref{sec:relatedwork}-\ref{sec:conclusion}.

@@

Both for function types and $\Sigma$, the output universe level is given as the
maximum of the levels of the constituent types, e.g.\ $(x : A)\ra B : \Set_{\max(i,j)}$
when $A : \Set_i$ and $B : \Set_j$.

@@

\subsection{Cumulativity}
\label{sec:cumulativity}

@@

A \textbf{category with families} (CwF), where all four underlying sets (of
objects, morphisms, types and terms) are in $\Set_i$. Following notation in
\cite{kaposi2019constructing}, we denote these respectively as $\Con : \Set_i$,
$\Sub : \Con \ra \Con \ra \Set_i$, $\Ty : \Con \ra \Set_i$ and $\Tm : (\Gamma :
\Con) \ra \Ty\,\Gamma \ra \Set_i$. We use $\id$ and $\blank\circ\blank$ to
denote identity and composition for substitution. We denote the empty context as
$\emptycon : \Con$, and the unique substitution into the empty context as
$\epsilon : \Sub\,\Gamma\,\emptycon$. Context extension is $\blank\ext\blank :
(\Gamma : \Con)\ra \Ty\,\Gamma \ra \Con$. Substitution on types and terms is
written as $\blank[\blank]$. Projections are noted as $\p :
\Sub\,(\Gamma\,\ext\,A)\,\Gamma$ and $\q : \Tm\,(\Gamma\,\ext\,A)\,(A[\p])$, and
substitution extension is $\blank,\blank : (\sigma : \Sub\,\Gamma\,\Delta)\ra
\Tm\,\Gamma\,(A[\sigma])\ra \Sub\,\Gamma\,(\Delta\ext A)$.

@@

\textbf{External function space} $\Pie : (A : \Set_j)\ra(A \ra \Ty\,\Gamma)\ra
\Ty\,\Gamma$, with $\appe : \Tm\,\Gamma\,(\Pie\,A\,B)\ra((x : A)\ra
\Tm\,\Gamma\,(B\,x))$ and its inverse $\lame$.

\textbf{Infinitary function space} $\Piinf : (A : \Set_j)\ra(A \ra
\Tm\,\Gamma\,\U)\ra \Tm\,\Gamma\,\U$, with $\appinf :
\Tm\,\Gamma\,(\El\,(\Pie\,A\,b))\ra((x : A)\ra \Tm\,\Gamma\,(\El\,(b\,x)))$ and
its inverse $\laminf$.

@@

\begin{definition}[Notion of signature]\label{def:signature}
A QIIT signature at level $j$ is a context in an arbitrary $M : \ToS_{i,j}$
model. We define the type of such signatures as follows:
\[
  \Sig_j :\equiv (i : \mathbb{N})\ra(M : \ToS_{i,j})\ra \Con_M
\]

Here, $j$ refers to the level of external types appearing in the signature, in
the domains of $\Pie$ and $\Piinf$ functions, while the quantified $i$ level is
required to allow interpreting a signature in arbitrary-sized ToS models. Note
that $\Sig_j$ is universe-polymorphic, so it is a family of cETT types and it is
not in any cETT universe.

\begin{example}{
    Signature for natural numbers. Here, no external types appear, so the level
    can be chosen as $0$.}
\begin{alignat*}{3}
  & NatSig : \Sig_0 \\
  & NatSig :\equiv \lambda (i :\mathbb{N})(M : \ToS_{i,0}).\\
  & \hspace{1em}(\emptycon_M\,\ext_M\, (N : \U_M) \,\ext_M\,(zero : \El_M\,N)\\
  & \hspace{2.67em}\ext_M (suc : N\arri_M\El_M\,N))
\end{alignat*}
\end{example}

With this, we are able to specify QIITs, and we can also interpret each
signature in an arbitrary ToS model, by applying a signature to a model.
$\Sig_j$ can be viewed as a precursor to a Church-encoding for the theory of
signatures, but we only need contexts encoded in this way, and not other ToS
components. In functional programming, this is sometimes called ``finally
tagless'' \cite{carette2007finally}, and it is used for defining and interpreting
embedded languages.
\end{definition}

@@

\begin{example}{Recursive equations}. Again, the universe is closed under $\Id$, which allows us to write equations
in $\Pii$ domains. A minimal (and trivial) example:
\begin{alignat*}{3}
  & RecEqSig :\equiv\\
  & \hspace{0.5em}\emptycon\,\ext (A : \U) \ext (a : \El\,A)\ext (f : \Pii\,(x : A)\,(\Id\,A\,x\,a\arri\El\,A))
\end{alignat*}
More interesting (and complicated) examples for recursive equations are boundary
conditions in various cubical type theories \cite{cohen2016cubical, angiuli2016computational, angiuli2018cartesian}. Note that our $\Id$
allows iterated equations as well, but these are all trivial in the semantics,
where we assume UIP.
\end{example}

\emph{Remark.} Since signatures are parametrized by a single universe level, all
external types in constructors must be contained in the same $\Set_j$
universe. We opted for this setup for the sake of simplicity. Cumulativity helps
here: it allows us to pick a $j$ level which is large enough to accommodate all
external types in a signature.

@@

In the following, for given $i$ and $j$ levels, we define a model $\bM_{i,j} :
\ToS_{\max(i+1,j)+1, j}$ such that $\Con_{\bM_{i,j}}$ is a type of structured
categories (of algebras). The level $i$ marks the level of all internal sorts in
an algebra, and the level $j$ marks the level of all external sets in function
domains. Hence, every algebra has level $\max(i+1,j)$. The bump is only needed
for $i$, since algebras merely contain elements of $A : \Set_j$ types, while
inductive sets are themselves elements of $\Set_i$. For example, $NatAlg_i :
\Set_{\max(i+1,0)} : \Set_{\max(i+1, 0)+1}$.

@@

We define $\bCon : \Set_{\max(i+1,j)+1}$ as $\flCwF_{\max(i+1,j)}$.
%% We define
%% $\flCwF$ as follows.

\begin{definition}[Finite limit CwFs]\label{def:flCwF}
For each level $i$ we define $\flCwF_i : \Set_{i+1}$ as an iterated $\Sigma$-type
with the following components:
\begin{enumerate}
  \item A CwF with underlying sets all in $\Set_i$. We reuse the component
    notations from Definition \ref{def:tos}.
  \item $\Sigma$-type $\Sg : (A : \Ty\,\Gamma)\ra \Ty\,(\Gamma \ext A)
    \ra \Ty\,\Gamma$, with term formers $\proj1$, $\proj2$ and $\blank,\blank$.
  \item Identity type $\Id : (A : \Ty\,\Gamma)\ra \Tm\,\Gamma\,A\ra
    \Tm\,\Gamma\,A\ra \Ty\,\Gamma$, with $\refl$, equality reflection and UIP.
  \item Constant families. This includes a type former $\Kfam : \Con \ra
    \Ty\,\Gamma$, where $\Gamma$ is implicitly quantified, together with $\lamK
    : \Sub\,\Gamma\,\Delta \ra \Tm\,\Gamma\,(\Kfam\,\Delta)$ and its inverse
    $\appK$. The idea is that $\Kfam\,\Delta$ is a representation of $\Delta$ as
    a type in any context. Clairambault and Dybjer called constant families
    ``democracy'' in \cite{clairambault2014biequivalence}.
\end{enumerate}
\end{definition}

@@

\begin{theorem}[Equivalence of initiality and induction, c.f. \cite{kaposi2019constructing}]
An object $\Gamma : \Con_{\bGamma}$ supports induction if and only if it is
initial. Moreover, induction and initiality are both proof irrelevant
predicates. \qed
\end{theorem}

@@

\subsection{Substitutions}
\label{sec:substitutions}

A \emph{weak flCwF morphism} $\boldsymbol{\sigma : \Sub\,\Gamma\,\Delta}$ is a functor between underlying categories, which additionally maps types to types and terms to terms, and satisfies the following:
  \begin{enumerate}
    \item $\bsigma\,(A[\sigma]) = (\bsigma\,A)\,[\bsigma\,\sigma]$
    \item $\bsigma\,(t[\sigma]) = (\bsigma\,t)\,[\bsigma\,\sigma]$
    \item The unique map $\epsilon : \Sub\,(\bsigma\,\emptycon)\,\emptycon$ has a retraction.
    \item Each $(\bsigma\,\p,\,\bsigma\,\q) : \Sub\,(\bsigma\,(\Gamma\,\ext\,A))\,(\bsigma\,\Gamma\,\ext\,\bsigma\,A)$ has an inverse.
  \end{enumerate}

@@

Note that the definition we just gave lives in $\Set_{\max(i+1,j)}$, but by
cumulativity it is also in $\Set_{\max(i+1,j)+1}$, as required by our $\bM_{i,j}
: \ToS_{\max(i+1,j)+1, j}$ specification of the model being defined.

@@

We define $\boldsymbol{\Ty\,\Gamma} : \Set_{\max(i+1, j)+1}$ as the type of
split flCwF-isofibrations over $\bGamma$, at level $\max(i+1, j)$. We extend
Ahrens' and Lumsdaine's displayed categories and their definition of
isofibrations \cite{displayedcats}.  We first define displayed flCwFs, then
specify iso-cleaving as additional structure on top of that.

\begin{definition}[Displayed flCwF] The type of displayed flCwFs at level $i$ is given as the logical predicate interpretation (see e.g.\ \cite{bernardy12parametricity} or \cite{hiit}) of $\flCwF_i$. For each flCwF component in $\bGamma$, there is a component in a displayed flCwF which ``lies over'' it.

\emph{Notation.} In situations where we need to refer to both ``base'' and
displayed things, we give \ul{underlined} names to contexts, substitutions,
types and terms in a base flCwF. For example, we may have $\ulGamma :
\Con_{\bGamma}$ living in $\boldsymbol{\Gamma : \Con}$, and $\Gamma :
\Con_{\bA}\,\ulGamma$ living in a displayed flCwF over $\bGamma$. We only use
underlining on cETT variable names, and overload flCwF component names for
displayed counterparts. For example, a $\Con$ component is named the same in
a base flCwF and a displayed one.

@@

In contrast to \cite{kaposi2019constructing}, it is not sufficient to model
types with displayed flCwFs. In ibid.\ substitutions are modeled as strict
morphisms, which we cannot do because our infinitary function types and identity
types cannot be interpreted strictly, as we will see in Section
\ref{sec:infinitaryfunction}. Strict preservation is expressed with metatheoretic equality, which
is respected by all cETT constructions, but isomorphisms are not necessarily
respected. Hence, we need to add additional structure to displayed flCwFs which
expresses preservation of base isomorphisms.

@@

\emph{Remark.} It is not possible to model types as fibrations or opfibrations,
because we have no restriction on the variance of ToS types. For example, the
type which extends a pointed set to a natural number signature, is neither a
fibration nor an opfibration.

@@

Iso-cleaving is given iso-cleaving in $\bA$ and the action of $\bsigma$ on
isomorphisms, e.g.\ we have $\coe_{\boldsymbol{A[\sigma]}}\,\ulsigma\,\Gamma
:\equiv \coe_{\bA}\,(\bsigma\,\ulsigma)\,\Gamma$.

Functoriality of type substitution, i.e.\ $\boldsymbol{A[\id]} = \bA$ and
$\boldsymbol{A[\sigma\circ\delta]} = \boldsymbol{A[\sigma][\delta]}$, follows
from Lemma \ref{lem:idcomppres} and split cleaving given by $\coe^{\id}$,
$\coe^{\circ}$, $\coh^{\id}$ and $\coh^{\circ}$ laws in $\bA$.

@@

$\boldsymbol{\Tm\,\Gamma\,A} : \Set_{\max(i+1, j)+1}$ is defined as the type of
\emph{weak flCwF sections} of $\bA$. The underlying functions of $\bt :
\bTm\,\bGamma\,\bA$ are as follows:
\begin{alignat*}{3}
  & \bt : (\ulGamma : \Con_{\bGamma}) \ra \Con_{\bA}\,\ulGamma\\
  & \bt : (\ulsigma : \Sub_{\bGamma}\,\ulGamma\,\ulDelta)
         \ra \Sub_{\bA}\,(\bt\,\ulGamma)\,(\bt\,\ulDelta)\,\ulsigma\\
  & \bt : (\ulA : \Ty_{\bGamma}) \ra \Ty_{\bA}\,(\bt\,\ulGamma)\,\ulA\\
  & \bt : (\ult : \Tm_{\bGamma}\,\ulGamma\,\ulA) \ra
          \Tm_{\bA}\,(\bt\,\ulGamma)\,(\bt\,\ulA)\,\ult
\end{alignat*}
Such that
\begin{enumerate}
  \item $\bt\,(\ulA[\ulsigma]) = (\bt\,\ulA)\,[\bt\,\ulsigma]$
  \item $\bt\,(\ult[\ulsigma]) = (\bt\,\ult)\,[\bt\,\ulsigma]$
  \item The unique map $\epsilon_{\bA} : \Sub\,(\bt\,\emptycon)\,\emptycon\,\id$ has a vertical retraction.
  \item Each $(\bt\,\p,\,\bt\,\q) : \Sub\,(\bt\,(\ulGamma\,\ext\,\ulA))\,(\bt\,\ulGamma\,\ext\,\bt\,\ulA)\,\id$ has a vertical inverse.
\end{enumerate}

Similarly to Section \ref{sec:substitutions}, we denote the evident preservation
isomorphisms as $\bt_{\emptycon} : \bt\,\emptycon \simeq_{\id} \emptycon$ and
$\bt_{\ext} : \bt\,(\ulGamma\ext \ulA) \simeq_{\id} \bt\,\ulGamma \ext
\bt\,\ulA$. In short, weak section is a dependently typed analogue of weak
morphism, with dependent underlying functions and displayed preservation
isomorphisms. We also have the derived fl-preservation isomorphisms.

@@

We define $\boldsymbol{\U : \Ty\,\Gamma}$ as the isofibration which is
constantly $\bSet_i$. A constant isofibration does not actually depend on the
base flCwF, and has trivial iso-cleaving where $\coe$-s are identity
functions. Hence, we have $\Con_{\bU}\,\ulGamma :\equiv \Set_i$ and
$\Sub_{\bU}\,\Gamma\,\Delta\,\ulsigma :\equiv \Gamma \ra \Delta$.

\emph{Remark.} The type $\bTm\,\bGamma\,\bU$ is strictly equal to
$\bSub\,\bGamma\,\bSet_i$, so it is helpful to think about semantic elements of
the universe as weak morphisms from $\bGamma$ to $\bSet_i$.

@@

For $\boldsymbol{a : \Tm\,\Gamma\,\U}$ and $\boldsymbol{B :
  \Ty\,(\Gamma\ext\El\,a)}$, we aim to define $\boldsymbol{\Pi\,a\,B}
\boldsymbol{:} \bTy\,\bGamma$. We define this as a dependent product of
isofibrations, indexed by a discrete domain. The discreteness is essential: with
a general $\bA \boldsymbol{:} \bTy\,\bGamma$ domain, $\bPii$ would not be
definable because of variance issues. Indeed, the category of categories is not
locally cartesian closed and does not support a general $\Pi$ type
\cite{johnstone2002sketches}.

@@

For $A : \Set_j$ and $\bB : A \ra \bTy\,\bGamma$, we define $\bPie\,A\,\bB
\boldsymbol{:} \bTy\,\bGamma$ as the $A$-indexed direct product of $\bB$. Since
the indexing is given by a metatheoretic function, every component is given in the
evident pointwise way.

\subsection{Infinitary Function Space}
\label{sec:infinitaryfunction}
